---
title: "Custom metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
Sys.setenv("_R_CHECK_CRAN_INCOMING_" = "true") # always build vignette like on CRAN
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we will demonstrate some advanced use cases of 
adding custom metadata to the NBDCtools enviornment, and use custom
metadata in various functions. 

## Prepare data

> **NOTE:** Before running the code in this vignette, we assume that you have
already read the [Get started](NBDCtools.html) vignette or the [Join
data](join.html) vignette, and know how to load the data for a given NBDC study
release.

To load `NBDCtools`, use the following command:

```{r setup}
library(NBDCtools)
suppressPackageStartupMessages({
  library(dplyr)
})
```

To demonstrate the custom metadata utilities, we load some trimmed ABCD
metadata files and simulated ABCD files that are included with the package.

metadata files:

```{r}
# assign the metadata objects to global and we will use later
purrr::walk(
  c("lst_dds", "lst_levels", "lst_sessions"),
  ~ {
    value <- readRDS(system.file(
      "extdata", "meta_internal", glue::glue("{.x}.rds"),
      package = "NBDCtools"
    ))[[1]][[1]]
    assign(gsub("lst_", "", .x), value = value, envir = sys.frame(which = 0))
  }
)
dds[1:5, 1:5]
levels[1:5, ]
sessions[1:5, ]
```

Simulated data files:
```{r}
dir_abcd <- system.file("extdata", "phenotype", package = "NBDCtools")
list.files(dir_abcd)
```


## Add custom metadata

For users who have their own metadata, they can add the metadata to the NBDCtools
by using the `add_custom_metadata()` function. This function will add the 
metadata to the NBDCtools environment, and make it available for use in various functions. 
The metadata should be can be used on `dd`, `levels`, and `sessions` tables, 
and should have the same format as the original metadata tables.

Now we can add previously loaded metadata to the NBDCtools environment:
```{r}
add_custom_metadata(
  dd = dds,
  levels = levels,
  sessions = sessions
)
```

## get custom metadata

To test if the custom metadata is added successfully, 
we can use the `get_<x>` function to retrieve the metadata from the 
NBDCtools environment. Please read [Metadata](metadata.html) vignette for more 
details on how to use the `get_<x>` functions.

The key of using custom metadata is to specify the **release** to be `"custom"`.

```{r}
get_dd_abcd(release = "custom")[1:5, 1:5]
```

When using the custom metadata, `study` argument is still needed but it can be 
either `"abcd"` or `"hbcd"`.

```{r}
get_dd_hbcd(release = "custom")[1:5, 1:5]
```

We can see they give the same `dd` table. 

## Use custom metadata in data preparation

The `release = "custom"` can be used in most functions that accept 
`release` argument.
Custom metadata can be very useful when preparing the data, especially for internal 
use when the pre-release metadata is created but not in the `NBDCtoolsData` package yet. 


```{r}
mydata <- create_dataset(
  study = "abcd",
  dir = dir_abcd,
  release = "custom",
  vars = c("mr_y_qc__raw__dmri__r01__series_t")
)
mydata
```


## Use case

One use case could be when one uses the release metadata, e.g. `get_get_dd_abcd()`,
and finds some errors in the metadata, they can fix the errors and add the 
fixed metadata as custom metadata, and use the custom metadata to prepare the data.

For example, the label of the dataset we just created for 
`mr_y_qc__raw__dmri__r01__series_t` is `Diffusion MRI (run 1): Series time`. 

```{r}
attr(mydata$mr_y_qc__raw__dmri__r01__series_t, "label")
```

It is stored in the `dd`

```{r}
get_dd_abcd(release = "custom") |> 
  filter(name == "mr_y_qc__raw__dmri__r01__series_t") |>
  select(name, label)
```

If we find the label is not correct, we can fix it in the `dd` table, and add the
fixed `dd` as custom metadata, and use the custom metadata to create the dataset again.

```{r}
# fix the label in the dd
dd_fixed <- get_dd_abcd(release = "custom") |>
  mutate(
    label = if_else(
      name == "mr_y_qc__raw__dmri__r01__series_t",
      "my custom label",
      label
    )
  )
# add the fixed dd as custom metadata
# we can leave levels and sessions empty so they will not be changed
add_custom_metadata(dd = dd_fixed)
# create the dataset again
mydata_fixed <- create_dataset(
  study = "abcd",
  dir = dir_abcd,
  release = "custom",
  vars = c("mr_y_qc__raw__dmri__r01__series_t")
)
attr(mydata_fixed$mr_y_qc__raw__dmri__r01__series_t, "label")
```
