---
title: "Filter data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
Sys.setenv("_R_CHECK_CRAN_INCOMING_" = "true") # always build vignette like on CRAN
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we will demonstrate how to filter a loaded dataset using
different conditions, such as by participant ID & session ID, event types,
excluding empty rows/columns, or combination of these.

## Prepare data

> **NOTE:** Before running the code in this vignette, we assume that you have
already read the [Get started](NBDCtools.html) page or the [Join
data](join.html) vignette, and know how to load data for a given NBDC study
release.

For demonstration purposes, we load a simulated ABCD dataset that is included
with the package.

```{r}
library(NBDCtools)

data <- readRDS(
  system.file("extdata", "simulated_data_abcd.rds", package = "NBDCtools")
)

dplyr::glimpse(data)
```

## Filter by participant ID and session ID

The `filter_id_events()` function accepts a vector of concatenated participant
and session IDs, separated by an underscore `_`, or a data frame with
`participant_id` and `session_id` columns. The function will filter the data by
the specified participant/events and return a data frame with the filtered
data.

```{r}
# filter using a vector of concatenated participant and session IDs
vec_id_events = c(
  "sub-0000000006_ses-04A", 
  "sub-0000000007_ses-05A",
  "sub-0000000002_ses-03A"
)
data |> 
  filter_id_events(vec_id_events)

# filter using a data frame with participant_id and session_id columns
data_filter <- dplyr::tribble(
  ~participant_id,  ~session_id,
  "sub-0000000006", "ses-04A", 
  "sub-0000000007", "ses-05A", 
  "sub-0000000002", "ses-03A"
)
data |>
  filter_id_events(data_filter)
```

Alternatively, if we only want to keep the rows that *do not match* the
specified participant and session IDs, we can set the `revert = TRUE` argument.

```{r}
data |> 
  filter_id_events(vec_id_events, revert = TRUE)
```

## Filter ABCD events

The `filter_events_abcd()` function provides a convenient way to specify
event-based filters for ABCD datasets using shorthands. These shorthands can be
a single condition, such as `"annual"`, `"substudy"`, or `"odd"`, or a
combination of conditions provided as a vector of strings, like `c("annual",
"even")`.

The following conditions are available:

- `"core"`: events for the ABCD core study
- `"annual"`: annual events for the ABCD core study
- `"mid_year"`: mid-year events for the ABCD core study
- `"substudy"`: events for ABCD substudies
- `"covid"`: events for the COVID substudy
- `"sdev"`: events for the Social Development substudy
- `"even"`: even-numbered events
- `"odd"`: odd-numbered events
- numerical expressions like `>2` or `<=5` to filter events by number
- any other string to be used as filter for the `session_id` column

```{r}
# retain only annual events
data |> 
  filter_events_abcd("annual")

# retain only even annual events ("imaging events")
data |> 
  filter_events_abcd(c("annual", "even"))

# retain only odd annual events before the 4-year follow-up
data |> 
  filter_events_abcd(c("annual", "odd", "<4"))

# retain only screener events (using a string condition to filter events)
data |> 
  filter_events_abcd(c("00S"))
```

## Filter empty rows

When combining data from different tables, we can end up with rows that do not
have values for any of the variables (besides the identifier columns). We
typically do not want to keep those rows as they don't contain any useful
information. The package provides a function to remove such
rows—`filter_empty_rows()`.

The simulated dataset does not have empty rows. To demonstrate the function, we
can manually set some rows to `NA` values and apply the `filter_empty_rows()`
function afterwards.

```{r}
data_empty <- data[1:5, ] |>
  dplyr::mutate(dplyr::across(-c(participant_id, session_id), ~ NA)) |> 
  dplyr::bind_rows(data[-(1:5), ])
data_empty

# filter empty rows
data_empty |> 
  filter_empty_rows()
```

As you can see, the first 5 rows that had missing values for all variables were
removed.

> **NOTE:**</span> The `filter_empty_rows()` function is automatically called by
the `join_tabulated()` or `create_dataset()` functions when joining data from
NBDC study datasets, so you do not need to call it explicitly. If you want to
keep the empty rows, you can set the `remove_empty_rows = FALSE` argument in
these two functions.

## Filter empty columns

Similarly, datasets can also contain completely empty columns. We can remove
them from the data using the `filter_empty_cols()` function.

Here, we modify the simulated data to create two empty columns and execute
`filter_empty_cols()` to remove them.

```{r}
data_empty_cols <- data |> 
  dplyr::mutate(
    ab_g_dyn__visit_type = NA,
    ab_g_dyn__cohort_grade = NA
  ) 
data_empty_cols

# filter empty columns
data_empty_cols |> 
  filter_empty_cols()
```

## Combining filters

We can combine several of the above filters by chaining the functions using the
pipe operator. For example, we can first exclude certain participant/events,
then filter by events, and finally remove empty rows using the following code:

```{r}
data_empty |> 
  filter_id_events(
    id_events = c("sub-0000000002_ses-03A"),
    revert = TRUE
  ) |>
  filter_events_abcd("odd") |>
  filter_empty_rows()
```
