---
title: "Join data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
Sys.setenv("_R_CHECK_CRAN_INCOMING_" = "true") # always build vignette like on CRAN
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we will demonstrate how to load/join selected variables and/or
tables from data files downloaded from the NBDC Data Hub into a single data
frame in memory.

```{r}
library(NBDCtools)
```

To download data from the ABCD or HBCD studies through the NBDC Data Hub, users
need to have a valid Data Use Certification (DUC) (see
[here](https://www.nbdc-datahub.org/data-access-process) for more details). In
this documentation, we will use simulated ABCD data files to demonstrate the
package's functionality.

You can use the following command to inspect the simulated data files:

```{r}
dir_abcd <- system.file("extdata", "phenotype", package = "NBDCtools")
list.files(dir_abcd)
```

We can see that the `phenotype/` directory contains several files. Each file
contains the data from a database table in the ABCD study. For example, the
`ab_g_dyn.parquet` file contains variables from the
`r NBDCtools:::get_url_deap_table_abcd("ab_g_dyn", link_name = "label")` table.

In a more realistic scenario, once you downloaded the ABCD data files to a local
directory, there will be a lot more files in the `phenotype/` directory; it
should contain one file per database table for a given study. To see the tables
available in the ABCD tabulated data resource, you can use `NBDCtools`'
`get_dd_abcd()` function:

```{r}
dd_abcd <- get_dd_abcd()
dd_abcd |>
  dplyr::distinct(table_name, table_label)
```

## Join data

We can use the `join_tabulated()` function to load the data from the files in
the `phenotype/` directory. The function will read and join the data from
different tables based on the variables and/or tables specified in the
these four arguments:

- `vars`: individual variables of interest
- `tables`: full tables of interest
- `vars_add`: additional individual variables
- `tables_add`: additional full tables

The `vars` and `tables` arguments are used to specify the main variables and/or
tables of interest. They will be extracted from the respective files
on disk and will be joined by the identifier columns into a single data frame in
memory. The used join operation is a `full_join()`, i.e., the resulting data
frame retains all rows that have at least one non-missing value in the
chosen variables/tables.

The `vars_add` and `tables_add` arguments are used to specify additional
variables and/or tables to be joined with the main variables/tables of interest.
The used join operation for adding these variables is a `left_join()`, i.e.,
only values for already existing rows will be added and no new rows will be
created. This is useful for adding variables to the dataset that are important
for a given analysis but are not the main variables of interest (e.g.,
design/nesting or demographic information). By left-joining these variables, one
avoids creating new rows that contain only missing values for the main variables
of interest selected using `vars` and `tables`.

```{r echo=FALSE}
dd_internal <- arrow::read_parquet(
  file = file.path(dir_abcd, "../dd.parquet")
) 
vars <- dd_internal$name
tables <- dd_internal$table_name |> 
  unique()
```

```{r}
join_tabulated(
  dir_data = dir_abcd,
  study = "abcd",
  vars = c(
    "mr_y_qc__raw__dmri__r01__series_t",
    "ab_g_dyn__visit_type",
    "ab_g_dyn__cohort_grade", 
    "ab_g_dyn__visit__day1_dt",
    "ab_g_stc__gen_pc__01", 
    "ab_g_dyn__visit_age",
    "ab_g_dyn__visit_days", 
    "ab_g_dyn__visit_dtt"
  )
)
```

```{r echo=FALSE}
# cat("ℹ Joining 8 variables from 3 tables...")
# data <- readRDS(file.path(dir_abcd, "../simulated_data_abcd.rds"))
```

> **NOTE:** This is a simulated dataset, so the number of rows are significantly
smaller than the real data.

***`remove_empty_rows` argument***

Joining selected columns from different tables using `join_tabulated()` may
result in rows with only missing values. This can occur when some
participant/events lack data for the selected variables but have data for other,
unselected variables in the joined tables. Since rows without any data are not
useful for analysis, the `join_tabulated()` function includes the
`remove_empty_rows` argument to eliminate these rows from the resulting dataset.
By default, this argument is set to `TRUE`, meaning rows with all missing data
will be removed. To keep these rows, set the argument to `FALSE`.

### Shadow matrices

For HBCD data, the `join_tabulated()` function can also join shadow matrices,
which store the reasons for missing values. Shadow matrix files are located in
the same directory as the data files, with one `{table_name}_shadow.parquet`
file accompanying each `{table_name}.parquet` file. To load the shadow matrices,
use the `shadow = TRUE` argument in the `join_tabulated()` function. This will
load selected columns or tables from the shadow matrix files and combine them
into a single table, just like the main data. For more information about shadow
matrices, see the [Work with shadow matrices](shadow.html) vignette.

```{r eval=FALSE}
join_tabulated(
  dir_data = dir_abcd,
  study = "hbcd",
  vars = ...,
  tables = ...,
  ...,
  shadow = TRUE
)
```
