---
title: "Work with metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
Sys.setenv("_R_CHECK_CRAN_INCOMING_" = "true") # always build vignette like on CRAN
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we will demonstrate how to retrieve and work with metadata for
NBDC studies, including data dictionaries, levels tables, session/event
information, and identifier columns.

## Overview

The `NBDCtools` package provides functions to access the following metadata
elements for ABCD and HBCD studies:

- **Data dictionary**: Variable and table definitions, data types, and other
  information.
- **Levels table**: Value, labels, and order for categorical variables.
- **Sessions table**: Information about study sessions/events.
- **Identifier columns**: Variables used to identify unique observations.

## Setup

```{r}
library(NBDCtools)
```

## Data dictionary

The data dictionary contains information about all variables in the tabulated
data for a given study, including their names, labels, data types, which
tables they belong to, etc. To read more about the structure of the data
dictionary for NBDC studies, see
[here](https://docs.abcdstudy.org/latest/documentation/curation/metadata.html#dd_levels_tables).

### Basic usage

```{r error = TRUE}
# Get data dictionary for the ABCD Study (latest release)
get_dd("abcd")

# Get data dictionary for the HBCD Study (latest release)
get_dd("hbcd")

# Get data dictionary for a specific release
get_dd("abcd", release = "6.0")

# If you are not sure what releases are available, just use a random number
# and the function will return an error message presenting the available
# releases
get_dd("abcd", release = "999.0")
```

### Study-specific functions

For convenience, you can use study-specific functions that do not require
specifying the study parameter:

```{r}
# ABCD-specific function
get_dd_abcd()

# HBCD-specific function  
get_dd_hbcd(release = "1.0")
```

### Filtering by variables

You can retrieve a subset of the data dictionary for specific variables:

```{r}
# Get data dictionary for specific variables
vars_of_interest <- c(
  "ab_g_dyn__visit_dtt", 
  "ab_g_dyn__visit_age", 
  "ab_g_stc__design_id__fam"
)
get_dd("abcd", vars = vars_of_interest)
```

### Filtering by tables

You can also retrieve a subset of the data dictionary for specific tables:

```{r}
# Get data dictionary for specific tables
tables_of_interest <- c(
  "ab_g_dyn",
  "ab_g_stc"
)
get_dd_abcd(tables = tables_of_interest)
```

## Levels table

The levels table provides value labels for categorical variables, showing what
each numeric value in the data corresponds to, as well as the order of levels.
To read more about the structure of the levels table for NBDC studies, see
[here](https://docs.abcdstudy.org/latest/documentation/curation/metadata.html#dd_levels_tables).

### Basic usage

```{r}
# Get levels table for ABCD Study (latest release)
get_levels("abcd")

# Get levels table for HBCD Study (latest release)
get_levels("hbcd")

# Get levels table for a specific release
get_levels("abcd", release = "6.0")
```

### Study-specific functions

```{r}
# ABCD-specific function
get_levels_abcd(release = "6.0")

# HBCD-specific function
get_levels_hbcd()
```

### Filtering by variables and/or tables

As for the data dictionary, you can also retrieve a subset of the levels table
for specific variables or tables:

```{r}
# Get levels for specific categorical variables
get_levels("abcd", vars = c("ab_g_dyn__visit_type"))

# Get levels for all categorical variables in specific tables
get_levels("abcd", tables = "ab_g_dyn")

# Get levels for a combination of specific variables and tables
get_levels_abcd(vars = "ab_g_dyn__visit_type", tables = "ab_g_stc")
```

## Sessions table

The sessions table contains information about the events/session IDs that are
part of a given release as well as their labels.

### Basic usage

```{r}
# Get sessions information for ABCD Study (latest release)
get_sessions("abcd")

# Get sessions information for HBCD Study (latest release)
get_sessions("hbcd")
```

### Study-specific functions

```{r}
# ABCD-specific function (for a specified release)
get_sessions_abcd(release = "6.0")

# HBCD-specific function (for a specified release)
get_sessions_hbcd(release = "1.0")
```

## Identifier columns

Identifier columns are the variables used to uniquely identify observations in
the dataset. These columns are essential for joining data from different tables.
The `get_id_cols()` function retrieves the identifier columns for a given study.

### Basic usage

```{r}
# Get identifier columns for ABCD Study (latest release)
get_id_cols("abcd")

# Get identifier columns for HBCD Study (latest release)
get_id_cols("hbcd")
```

### Study-specific functions

```{r}
# ABCD-specific function (for a specified release)
get_id_cols_abcd(release = "6.0")

# HBCD-specific function (for a specified release)
get_id_cols_hbcd(release = "1.0")
```

## General metadata function

The `get_metadata()` function is the low-level function that is used by all
specific metadata functions. You can use it directly to retrieve any type of
metadata:

```{r}
# Get data dictionary (same as get_dd)
get_metadata("abcd", type = "dd", release = "6.0")

# Get levels table (same as get_levels)  
get_metadata("abcd", type = "levels", vars = "ab_g_dyn__visit_type")

# Get sessions table (same as get_sessions)
get_metadata("abcd", type = "sessions")
```
