---
title: "Work with shadow matrices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
Sys.setenv("_R_CHECK_CRAN_INCOMING_" = "true") # always build vignette like on CRAN
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we will demonstrate how to work with shadow matrices for NBDC
studies. Shadow matrices are used to store information about the reasons for
missing values in the main data (see
[here](https://docs.hbcdstudy.org/latest/datacuration/phenotypes/#shadow-matrices)
for more details). Currently, the HBCD Study releases shadow matrices, while the
ABCD Study does not (see below for how to use the R package `naniar` to create a
basic shadow matrix for ABCD data).

## Prepare data

> **NOTE:** Before running the code in this vignette, we assume that you have
already read the [Get started](NBDCtools.html) page or the [Join
data](join.html) vignette, and know how to load the data/shadow from a given
NBDC study release.

For demonstration purposes, we load a simulated HBCD dataset and shadow matrix
that are included with the package.

```{r}
library(NBDCtools)
data <- readRDS(
  system.file("extdata", "simulated_data_hbcd.rds", package = "NBDCtools")
)
dplyr::glimpse(data)

shadow <- readRDS(
  system.file("extdata", "simulated_data_hbcd_shadow.rds", package = "NBDCtools")
)
dplyr::glimpse(shadow)
```

> **IMPORTANT:** Before running any shadow-related functions, it is very
important to ensure that the *dimensions of the main data and shadow matrix
match*, i.e., that they have the same number of rows and columns, all the
columns in the main data are also present in the shadow matrix, and all the
identifier pairs (`participant_id`, `session_id`, `run_id`) exist in both data
frames.

The `join_tabulated()` function has a parameter `remove_empty_rows` that removes
rows that only have missing values by default. This can lead to a situation
where the main data and shadow matrix have different dimensions, which will
cause problems in the subsequent shadow-related functions. If you know that
downstream processing involves shadow matrices, make sure to set this parameter
to `remove_empty_rows = FALSE` when calling `join_tabulated()`.

This is not a problem if you are using `create_dataset()` function, as it
automatically sets `remove_empty_rows = FALSE` when creating a dataset with
shadow matrix.

## Bind shadow matrix to main data

### Using a provided shadow matrix

If you have a data frame with a shadow matrix, you can bind it to the main data
using the `shadow_bind_data()` function. This function will bind the shadow
matrix to the data using the identifier columns (`participant_id`, `session_id`,
and `run_id` by default). The appended shadow columns will have the same column
names as the corresponding columns in the main data, but with a suffix
(`_shadow` by default).

- If there are rows with identifier pairs in the main data that don't exist in
  the shadow matrix, the appended shadow matrix rows will be filled with `NA`
  values.
- If there are rows with identifier pairs in the shadow matrix that don't exist
  in the main data, these rows will be dropped from the resulting data frame.

```{r}
shadow_bind_data(
  data = data,
  shadow = shadow
) |> 
  dplyr::glimpse()
```

### Using a `naniar` shadow matrix

If there is no shadow matrix provided, like for ABCD data, or you do not wish to
use the provided shadow matrix, you can create a shadow matrix using the
[`naniar`](https://naniar.njtierney.com/) R package. To do so, set
`naniar_shadow = TRUE`.

> **IMPORTANT:** `naniar` is not a dependency of the `NBDCtools` package, so you
need to install it separately. You can do so using the following command:

```{r}
if (!requireNamespace("naniar", quietly = TRUE)) {
  install.packages("naniar")
}
```

To create a shadow matrix using `naniar`, you can use the following command:

```{r}
shadow_bind_data(
  data = data,
  shadow = NULL, # no shadow matrix provided
  naniar_shadow = TRUE
) |> 
  dplyr::glimpse()
```

The downside of this approach is that the shadow matrix will not contain
reasons for missing values, but will only indicate that the value is
missing. The benefit of using `naniar_shadow = TRUE` is that you can use
functionality from the `naniar` R package to explore and visualize missingness
in the data (see [here](https://naniar.njtierney.com/articles/naniar.html) for
more details).

## Fix missing values in the shadow matrix

Joining shadow matrices for different tables using `join_tabulated()` can result
in missing values in the shadow matrix for certain observations (e.g., because
one instrument was administered at more time points than another, or because
a participant did not complete a certain instrument). These missing values in
the shadow matrix could be mistakenly interpreted as indicating that the main
data has values for these cells. But indeed these cells are missing in both
the main data and the shadow matrix.

Here is a simplified example of this scenario:

```{r}
my_table1 <- dplyr::tibble(
  participant_id = c("sub-001", "sub-002"),
  session_id = c("ses-001", "ses-002"),
  run_id = c("run-001", "run-002"),
  var1 = c("reason1", NA),
  var2 = c(NA, "reason2")
)
my_table1

my_table2 <- dplyr::tibble(
  participant_id = c("sub-001", "sub-003"),
  session_id = c("ses-001", "ses-003"),
  run_id = c("run-001", "run-003"),
  var3 = c(NA, "reason3"),
  var4 = c("reason4", NA)
)
my_table2
```

When binding these two tables together, we will get the following result:

```{r}
id_table <- dplyr::full_join(
  dplyr::select(my_table1, participant_id, session_id, run_id),
  dplyr::select(my_table2, participant_id, session_id, run_id)
)
dplyr::left_join(id_table, my_table1) |> 
  dplyr::left_join(my_table2) 
```

We can see that there are missing values (`NA`) in the `var3` and `var4` columns
for the `sub-002`/`ses-002`/`run-002` row and in the `var1` and `var2` columns
for the `sub-003`/`ses-003`/`run-003` row. If this table was a shadow matrix,
this would indicate that there should be values for these cells in the main
data, but in reality the shadow matrix just has missing values due to joining
the two tables.

To fix this issue, you can use the `shadow_replace_binding_missing()`
function, which will replace the `NA` values in the shadow matrix with a custom
value (default `"Missing due to joining"`).

The simulated shadow matrix does not have cases where the shadow matrix has `NA`
values due to joining. For demonstration, we will manually convert a few values
to `NA`:

```{r}
shadow$mh_cg_ibqr_surg_001
# set them to NA
shadow$mh_cg_ibqr_surg_001[1:5] <- NA_character_
shadow$mh_cg_ibqr_surg_001
```

Now we can use the `shadow_replace_binding_missing()` function to replace
the `NA` values in the shadow matrix with an indicator: 

```{r}
shadow_replace_binding_missing(
  data = data,
  shadow = shadow
)
```

As you can see, the `mh_cg_ibqr_surg_001` column now has
`"Missing due to joining"` values.

> **IMPORTANT:** Unlike the `shadow_bind_data()` where unmatched rows between
the main data and the shadow matrix are allowed, this function requires the
dimensions of the data and shadow matrix to match, i.e., to have the same 
number of rows and columns. If they do not match, the function will throw an
error.
