---
title: "Transform data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
Sys.setenv("_R_CHECK_CRAN_INCOMING_" = "true") # always build vignette like on CRAN
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we will demonstrate how to apply different transformations to
a joined dataset such as converting categorical columns to factors, adding
variable and value labels, converting categorical values to labels, and more.

## Prepare data

> **NOTE:** Before running the code in this vignette, we assume that you have
already read the [Get started](NBCtools.html) vignette or the [Join
data](join.html) vignette, and know how to load the data for a given NBDC study
release.

To demonstrate the transformation functions, we load a simulated ABCD dataset
that is included with the package.

```{r}
library(NBDCtools)
data <- readRDS(
  system.file("extdata", "simulated_data_abcd.rds", package = "NBDCtools")
)
dplyr::glimpse(data)
```

## Convert categorical columns to factors

From looking at the simulated data, we can see that categorical columns like
`ab_g_dyn__visit_type` are of type "character". We can use `transf_factor()`
function to convert these columns to type "factor" which is the correct type
for categorical variables in R.

```{r}
data_transf <- data |>
  transf_factor(study = "abcd")
dplyr::glimpse(data_transf)
```

The `transf_factor()` function automatically detects categorical columns and
converts them to ordered or unordered factors based on the specification in the
data dictionary and levels table for a given study.

## Apply variable and value labels

Next, we can add variable and value labels to the dataset.

```{r}
data_transf <- data_transf |> 
  transf_label(study = "abcd")
```

To inspect the *variable labels*, we can use `sjlabelled::get_label()` function:

```{r}
sjlabelled::get_label(data_transf)
```

To inspect the *value labels*, we can use `sjlabelled::get_labels()` function:

```{r}
sjlabelled::get_labels(data_transf, attr.only = TRUE, values = "n")
```

If the labeling is done incorrectly, we can simply rerun the `transf_label()`
function to fix it. If we want to remove all labels, we can use

```{r}
data_labels_removed <- sjlabelled::remove_all_labels(data_transf) 
sjlabelled::get_label(data_labels_removed)
sjlabelled::get_labels(data_labels_removed, attr.only = TRUE, values = "n")
```

## Convert time columns to hms format

Time columns in the dataset (e.g. `mr_y_qc__raw__dmri__r01__series_t`) are
formatted as character strings `"HH:MM:SS"` by default. If we want to convert
these columns into [`hms`](https://hms.tidyverse.org/reference/hms-package.html)
format, we can use the `transf_time_to_hms()` function:

```{r}
data_transf <- data_transf |> 
  transf_time_to_hms(study = "abcd")
dplyr::glimpse(data_transf)
```

As we can see, the column type is converted from `character` to `time` (`hms`)
class.

## Convert categorical column levels to labels

In some cases, such as for creating plots, it is useful to convert categorical
values to labels. We can use the `transf_value_to_label()` function to do so:

```{r}
data_transf |> 
  transf_value_to_label()
```

> **NOTE:** Before running this function, make sure that the data has been
transformed with the `transf_factor()` and `transf_label()` functions, so that
the variable and value labels are available.

## Convert missing codes to `NA`

In ABCD and HBCD Study datasets, some of the categorical columns use specific
codes to denote missingness/non-responses (e.g., `"999"` for "Don't know" or
`"777"` for "Decline to answer"). If we want to remove these values before
analysis, we can use the `transf_value_to_na()` function to convert these codes
to `NA`.

> **NOTE:** By default, this function converts all standard categorical
missingness codes (`"222"` through `"999"`) to `NA`. In the ABCD Study, these
codes are consistenly used throughout the whole dataset; in the HBCD Study,
however, columns may use different codes for non-responses or missing values.
Please refer to the data dictionary and levels table for the specific study to
see which codes to convert.

```{r}
data_transf |> 
  transf_value_to_na()
```

The simulated dataset does not contain categorical missingness codes. However,
this function provides a custom parameter `missing_codes` to specify the codes
that should be converted to `NA`. For example, if we want to convert the
categorical values `"1"` and `"2"` to `NA`, we can use:

```{r}
data_transf |> 
  transf_value_to_na(missing_codes = c("1", "2"))
```

In the `ab_g_dyn__visit_type` column, we can see that the values `"1"` and `"2"`
have been converted to `NA`.

The function has another parameter, `ignore_col_pattern`, that can be used to
ignore specific columns, so that they are exempt from the conversion. This
parameter accepts a regular expression pattern, meaning all columns that match
the pattern will be ignored. For example, we can ignore all columns that start
with `ab_g_dyn__visit` by using:

```{r}
data_transf |> 
  transf_value_to_na(ignore_col_pattern = "^ab_g_dyn__visit")
```
